---
title: "Class08"
author: "Erica Birkholz"
date: "February 6, 2019"
output: word_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Class 08 - Unsupervised learning
# K-means

```{r}
kmeans(x, centers=3, nstart=20)
```

```{r}
# Generate some example data for clustering
tmp <- c(rnorm(30,-3), rnorm(30,3))
x <- cbind(x=tmp, y=rev(tmp))
plot(x)
```

# cbind takes data and binds it into columns.
# y is the reverse of x

```{r}
km <- kmeans(x, centers=2, nstart=20)
km
```
# "ss" means sum of squares (euclidean distance calculation)
withinss - distance of component to center
totss - total distances to center

# Does not reveal structure in data, imposes a structure.

#Size of clusters
```{r}
km$size
```
```{r}
km$cluster
```

```{r}
plot(x, col=km$cluster)
points(km$centers, col="blue", pch=15, cex=3)
```

## Hierarchical clustering
```{r}
dist_matrix <- dist(x) 
hc <- hclust(dist(x))
```

```{r}
plot(hc)
abline(h=6, col="red")
grp2 <- cutree(hc, h=6)
```

# Reveals structure! Unlike kmean.

```{r}
plot(x, col=grp2)
```

```{r}
plot(hc)
abline(h=2.5, col="blue")
grp6 <- cutree(hc, h=2.5)
table(grp6)
```

```{r}
# Draws a dendrogram
plot(hc)
abline(h=6, col="red")
cutree(hc, k=3 ) # Cut into k grps if you don't know the height that will give you the groups you want.
```

```{r}
# Using different hierarchical clustering methods
d <- dist_matrix
hc.complete <- hclust(d, method="complete")
hc.average <- hclust(d, method="average")
hc.single <- hclust(d, method="single")
```

Made up overlapping data - more life like!

```{r}
# Step 1. Generate some example data for clustering
x <- rbind(
 matrix(rnorm(100, mean=0, sd = 0.3), ncol = 2), # c1
 matrix(rnorm(100, mean = 1, sd = 0.3), ncol = 2), # c2
 matrix(c(rnorm(50, mean = 1, sd = 0.3), # c3
 rnorm(50, mean = 0, sd = 0.3)), ncol = 2))
colnames(x) <- c("x", "y")
# Step 2. Plot the data without clustering
plot(x)
# Step 3. Generate colors for known clusters
# (just so we can compare to hclust results)
col <- as.factor( rep(c("c1","c2","c3"), each=50) )
plot(x, col=col)
```

```{r}
d_real <- dist(x)

clust_real <- hclust(d_real)

plot(clust_real)

grp1 <- cutree(clust_real, k=3)

plot(x, col=grp1)

table(grp1)
```

## PCA in R

```{r}
## You can also download this file from the class website!
mydata <- read.csv("https://tinyurl.com/expression-CSV",
 row.names=1)
head(mydata)
```

```{r}
## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE) 
summary(pca)
```

Make our first PCA plot

```{r}
dim(pca$x)
plot(pca$x[,1], pca$x[,2], xlab="PC1", ylab="PC2")
```

```{r}
## lets do PCA
pca <- prcomp(t(mydata), scale=TRUE)
## See what is returned by the prcomp() function
attributes(pca)
# $names
#[1] "sdev" "rotation" "center" "scale" "x"
#
# $class
#[1] "prcomp"
```

```{r}
pca <- prcomp(t(mydata), scale=TRUE)
## A basic PC1 vs PC2 2-D plot
plot(pca$x[,1], pca$x[,2])
## Precent variance is often more informative to look at
pca.var <- pca$sdev^2
pca.var.per <- round(pca.var/sum(pca.var)*100, 1)

pca.var.per
```

```{r}
barplot(pca.var.per, main="Scree Plot",
 xlab="Principal Component", ylab="Percent Variation")

```

```{r}
## A vector of colors for wt and ko samples
colvec <- as.factor( substr( colnames(mydata), 1, 2) )
plot(pca$x[,1], pca$x[,2], col=colvec, pch=16,
 xlab=paste0("PC1 (", pca.var.per[1], "%)"),
 ylab=paste0("PC2 (", pca.var.per[2], "%)")) 
```



### Hands on session

```{r}
x <- read.csv("UK_foods.csv")

dim(x)
```

```{r}
# Note how the minus indexing works
rownames(x) <- x[,1]
x <- x[,-1]
head(x)
```

```{r}
barplot(as.matrix(x), beside=T, col=rainbow(nrow(x)))
```

```{r}
pairs(x, col=rainbow(10), pch=16)
```

```{r}
# Use the prcomp() PCA function 
pca <- prcomp( t(x) )
summary(pca)
```

Make PCA plot
```{r}
plot(pca$x[,1], pca$x[,2])
text(pca$x[,1], pca$x[,2], labels=colnames(x))
```

