---
title: "Class 14: RNAseq Analysis"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Import countData and colData

```{r}
counts <- read.csv("data/airway_scaledcounts.csv", stringsAsFactors = FALSE)
metadata <-  read.csv("data/airway_metadata.csv", stringsAsFactors = FALSE)
```


How many genes are in the dataset?

```{r}
nrow(counts)
```

How many counts?
```{r}
ncol(counts)
```
8, because 1st column is gene names.

Check metadata to counts correspondence (column to row match).
```{r}
all(metadata$id == colnames(counts[-1]))
```

Extract control exps (columns of **counts** based on **metadata**)
```{r}
#Find the column names (ids) of the 'control' experiments
control.ids <- metadata$id[metadata[,"dex"] == "control"]
control.ids

#Use the ids to extract the control counts columns
control <- counts[,control.ids]
rownames(control) <- counts$ensgene
head(control)
```

```{r}
#Find the column names (ids) of the 'treated' experiments
treated.ids <- metadata$id[metadata[,"dex"] == "treated"]
treated.ids

#Use the ids to extract the treated counts columns
treated <- counts[,treated.ids]
rownames(treated) <- counts$ensgene
head(treated)
```


Find the mean counts value for each gene in contorl exps.
```{r}
control.mean <- rowSums(control)/nrow(control)
head( control.mean)
```


Find the mean counts value for each gene in treated exps.
```{r}
treated.mean <- rowSums(treated)/nrow(treated)
head( treated.mean)
```

```{r}
meancounts <- data.frame(control.mean, treated.mean)
colSums(meancounts)
```

#Plot!

```{r}
plot(meancounts$control.mean,treated.mean, xlab="control", ylab="treated")
```


Make a log-log plot.

```{r}
plot(meancounts$control.mean,treated.mean, log="xy", xlab="log control", ylab="log treated")
```

```{r}
meancounts$log2fc <- log2(meancounts[,"treated.mean"]/meancounts[,"control.mean"])
head(meancounts)
```

We need to filter out zero value entries.
```{r}
#meancounts[,1:2]

toy <- c(1, 2, 0, 4, 0)
which(toy  == 0)

newtoy <- cbind(toy, toy)
newtoy
newtoy == 0
which(newtoy == 0, arr.ind = T)
unique( which(newtoy==0, arr.ind=T)[,1])
```

```{r}
zero.vals <- which(meancounts[,1:2]==0, arr.ind=TRUE)
head(zero.vals)

to.rm <- unique(zero.vals[,1])
mycounts <- meancounts[-to.rm,]
head(mycounts)
```

A common threshold used for calling something differentially expressed is a log2(FoldChange) of greater than 2 or less than -2. Let’s filter the dataset both ways to see how many genes are up or down-regulated.

```{r}
up.ind <- mycounts$log2fc > 2
down.ind <- mycounts$log2fc < (-2)

sum(up.ind)
sum(down.ind)
```
250 genes are upregulated over a log2 fold change of 2.
367 genes are downregulated based on that threshold.

What are the upregulated genes?

```{r}
head( mycounts[up.ind,])
```

Annotate the genes so they are biologically relevant to us.

```{r}
anno <- read.csv("data/annotables_grch38.csv")
head(anno)
```

```{r}
dim(anno)
```

Let's try the **merge()** function to add the annotations to the data.

```{r}
#head(mycounts)
mycounts.anno <- merge(x=mycounts, y=anno, by.x="row.names", by.y="ensgene")
head(mycounts.anno)
```

```{r}
#biocLite("DESeq2")
#biocLite("AnnotationDbi")
#biocLite("org.Hs.eg.db")
```

```{r}
library("AnnotationDbi")
library("org.Hs.eg.db")
```

```{r}
columns(org.Hs.eg.db)
```


```{r}
mycounts$symbol <- mapIds(org.Hs.eg.db,
                     keys=row.names(mycounts),
                     column="SYMBOL",
                     keytype="ENSEMBL",
                     multiVals="first")
```

```{r}
library(DESeq2)
```

Build our DESeq2 Data Set
```{r}
dds <- DESeqDataSetFromMatrix(countData=counts, 
                              colData=metadata, 
                              design=~dex, 
                              tidy=TRUE)
dds
```

Run the DESeq Analysis
```{r}
dds <- DESeq(dds)
```

```{r}
res <- results(dds)
res
```


#Volcano Plot!

Plotting log2 Fold Change and subtracting adjusted p value (padj)
```{r}
plot(res$log2FoldChange, -log(res$padj))
```

```{r}
mycols <- rep("gray", nrow(res))
mycols[abs(res$log2FoldChange) > 2] = "blue"
subset.inds <- res$padj < 0.01 & abs(res$log2FoldChange) > 2
mycols[subset.inds] <- "magenta"

plot(res$log2FoldChange, -log(res$padj), col=mycols)
```












